Open-Set Domain Adaptation through Self-Supervision

### 2022-02-09 paper/main.tex
Added remaining chapter 4

Let's denote with $rot({\bf x}, i)$ the rotation of the sample image {\bf x} by $i\times 90$ degrees clockwise.
This is the self-supervised part of our proposed model as rotations $i \in [0, 3]$ can be randomly generated and then predicted.
To avoid situations where the objective orientation of a sample would be a too complicated task, even for a human being, we also feed in input to the model the un-rotated image features.
Alternatively, instead to have the model predict the rotation of a sample for any class, we also try using a different head for each one of the known classes $\mathcal{C_s}$, along with different loss functions.

@@

To separate samples we train the model on an enhanced version of $\mathcal{D}_s$, $\tilde{\mathcal{D}_s} = \{({\bf x}_i^s, \tilde{\bf x}_i^s, z)\}$ where $\tilde{\bf x}_i^s$ is the rotated version of ${\bf x}_i^s$ and $z$ is the rotation index.
The network is composed by an extractor $E$ and two heads, $R_1$ and $C_1$ in its standard form.
When using a multi-head rotation predictor, it is composed of $|\mathcal{C}_s|+1$ heads for the rotation task, and an additional one for the classification task.
The features of both the original and rotated image are used to predict the rotation as $\tilde{z}_s = softmax(R_1([E({\bf x}^s), E(\tilde{\bf x}^s)]))$ while classes are predicted solely on un-rotated features $\tilde{\bf y}^s = softmax(C_1(E({\bf x}^s)))$.
The model is training by minimizing an objective function defined as $\mathcal{L}_1 = \mathcal{L}_{C_1} + \mathcal{L}_{R_1}$.
Both heads use a Cross-Entropy loss function:

@@

We also try using a different loss function for the rotation classifier, which is:

\begin{equation}
  \mathcal{L}_{R_1} = -\alpha_1\sum_{i\in\mathcal{D}_s} {\bf y}_i^s \log \tilde{{\bf z}}_i^s+\lambda_1||{\bf v}_i^s-\gamma({\bf z}_i^s)||_2^2
\end{equation}
\label{eq:center_loss}

In both versions $\alpha_1$ is the weight of the rotation task, $v_i$ is the penultimate layer of the rotation classifier $R_1$ and $\gamma({\bf z}_i)$ is the centroid of the features of class of ${\bf v}_i$.

>>>>

Let's denote with $rot({\bf x}, i)$ the rotation of the sample image {\bf x} by $i\times 90$ degrees clockwise.
This is the self-supervised part of our proposed model as rotations $i \in [0, 3]$ can be randomly generated and then predicted.
To avoid situations where the objective orientation of a sample would be a too complicated task, even for a human being, we also feed in input to the model the un-rotated image features.
Alternatively, instead to have the model predict the rotation of a sample for any class, we also try using a different head for each one of the known classes $\mathcal{C}_s$, along with different loss functions.

@@

To separate samples we train the model on an enhanced version of $\mathcal{D}_s$, $\tilde{\mathcal{D}_s} = \{({\bf x}_i^s, {\bf \tilde x}_i^s, z_i^s)\}_{i=1}^{N_s}$ where ${\bf \tilde x}_i^s$ is the rotated version of ${\bf x}_i^s$ and $z_i^s$ is the rotation index associated to image $i$.
The network is composed by an extractor $E$ and two heads, $R_1$ and $C_1$ in its standard form.
When using a multi-head rotation predictor, it is composed of $|\mathcal{C}_s|+1$ heads for the rotation task, and an additional one for the classification task.
For the roation index prediction, we both use the single-head and multi-head architecture.
The features of both original and rotated image are used to predict the rotation as ${\bf \tilde z}_s = softmax(R_1([E({\bf x}^s), E({\bf\tilde x}^s)]))$ in the single-head case and using the $j$-th head as ${\bf\tilde{z}}_s = softmax(R_{1,j}[E({\bf x}^s), E({\bf\tilde{x}}^s)])$ in the multi-head case.
Classes are predicted only using un-rotated features as $\tilde{\bf y}^s = softmax(C_1(E({\bf x}^s)))$.
The model is training by minimizing an objective function defined as $\mathcal{L}_1 = \mathcal{L}_{C_1} + \mathcal{L}_{R_1}$.
This is the sum of two cross-entropy loss functions as in:

@@

Where $\alpha_1$ is a weight associated to the rotation task.
We also try using an extended rotation objective function $\mathcal{L}_{R_1}^*$ also implementing a center loss function\cite{CenterLoss}:

\begin{equation}
  \mathcal{L}_{R_1}^* = -\alpha_1\sum_{i\in\mathcal{D}_s} {\bf y}_i^s \log \tilde{{\bf z}}_i^s+\lambda_1||{\bf v}_i^s-\gamma({\bf z}_i^s)||_2^2
\end{equation}
\label{eq:center_loss}

Here {\bf $v_i$} is the penultimate layer of the rotation classifier, called {\it features}, and $\gamma({\bf z}_i)$ gives the center of the features ${\bf v}_i$ associated to class $i$ and $||\cdot||_2^2$ is the $l$-2 norm and $\lambda_1$ is the weight associated with this extension of the loss function.

When training is completed, we can start separating samples.
To do so, we get the normality score $\mathcal{N}(\cdot)$ which is defined as the maximum prediction of the rotation classifier, $\mathcal{N}({\bf\tilde{x}}_i^s) = \max({\bf\tilde{z}}_i^t)$.
To tell wheter a sample belongs to the known samples of the target domain $\mathcal{D}_t^{knw}$ or the unknown one $\mathcal{D}_t^{unk}$ requires choosing a threshold $\tilde{\mathcal{N}}$.
Then we can simply separate as:

\begin{equation}
  \begin{cases}
    {\bf\tilde x}_i^t \in \mathcal{D}_t^{knw} & \text{ if } \mathcal{N}({\bf\tilde x}_i^s) \geq \tilde{\mathcal{N}} \\
    {\bf\tilde x}_i^t \in \mathcal{D}_t^{unk} & \text{ if } \mathcal{N}({\bf\tilde x}_i^s) < \tilde{\mathcal{N}}
  \end{cases}
\end{equation}
\label{eq:sample_separation}

When employing a multi-head architecture, we need to choose which one of the $|\mathcal{C}_s|$ heads to use for the classification task.
Head $R_{1,j}$ is used by picking $j$ as $\displaystyle j = \arg\max_j {\bf\tilde{y}}_i^s$.

\subsection{Performance Metric}
\label{sec:performance_metric}

Evaluating model perfomance requires finding a balance between two metrics:
the first one is {\it OS*}, the share of correctly classified samples;
the second one is {\it UNK}, the share of correclty rejected samples.
One common problem is a model never confident enough to reject a sample as unknown, thus possibly achieving high {\it OS*} scores but near-zero {\it UNK}, and the opposite one, a model rejecting every given sample achieving perfect {\it UNK} but zero {\it OS*} scores.
To compare models in a robust manner, we picked an harmonic mean between {\it OS*} and {\it UNK}, defined as $HOS = 2\frac{OS^*\times UNK}{OS^*+UNK}$.
This type of mean is more biased towards the lowest of the two scores, resulting in a more severe evaluation of models.

\section{Experiments}
\label{sec:experiments}
